Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-29 ltx/data_paper/DataPaper.tex
commit for merge

Here we describe the problem of estimating non-linear state space neural field model using expectation-maximisation (EM) based algorithm \cite{Dempster1977,Shumway2000} is described. EM is an iterative algorithm that converges on the maximum-likelihood estimates of the parameters. Both the E and the M steps of the EM algorithm increase a lower bound on the log-likelihood function of the parameters given the observed field.  In the E-step we estimate the state $\mathbf x_t$, of the system given the observed field and current estimate of the model parameters to calculate expected log-likelihood. In the M-step we estimate the system parameters by maximising the expected log-likelihood computed in the E-step. The resulting parameters are then used to estimate the distribution of the state vector sequence for the next iteration. The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}.

@@

The estimated supports were then employed in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}\label{sec:EM-Discussion}
The correlation analysis specifies approximate supports for both the disturbance covariance function and connectivity kernel. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be set as described in \S\ref{subsec:CorrAnal}, the chosen values influence the exact shape of the kernel significantly, the analysis also assumes the sigmoidal firing rate in the active region, as a result  the kernel and disturbance amplitudes cannot be inferred only through the correlation analysis. This step might be considered as a prerequisite for the EM algorithm described in \S\ref{subsec:EM-EMAlgorithm} where the approximated supports are used to estimate the gain of the kernel and field disturbance variance. Knowing the spatial extent of the connectivity kernel greatly improves the speed of the estimation algorithm as the number of basis functions for the kernel decomposition is significantly reduced  compared to the case where the basis functions are arranged across the entire spatial domain. This also improves the uncertainty of the estimates as lower number of unknown parameters need to be estimated. The approximated support of the disturbance covariance function facilitates the application of the EM algorithm. This way the disturbance covariance matrix can be decomposed into two parts: an unknown scalar and a constant matrix depending only on the inferred spatial support. Therefore, the estimation problem of the disturbance covariance matrix breaks down to the estimation of a single scalar parameter, improving the accuracy and the convergence property of the algorithm.

The additional Taylor approximation step introduced in the E-step of the algorithm resembles the extended Kalman filtering \cite{Jazwinski1970,Chui2009}. EKF approximates a stationary non-linear dynamic system with a linear non-stationary one using first order Taylor expansion around the state estimates. This might imply that the extended Kalman smoother is a better candidate (in terms of implementation) for the state inference in the proposed algorithm. However, extended Kalman smoother (EKS) requires re-linearising about the backwards estimates which in turn necessitates the non-linear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation. In addition, URTSS provides better state approximation compared to the EKS \cite{Haykin2001} and therefore the constructed lower bound will be more accurate, resulting in a better parameter estimation.

While EKS has been  applied in the EM algorithm for learning non-linear models \cite{Ghahramani1999} this is the first work (to the best author's knowledge) where the URTSS is used in this context. However, as in the case of the EKS the stability of the algorithm cannot be ensured. This is because the use of EKS or URTSS represents an approximation to the exact E-step. As a result the maximisation of the approximated lower bound might not necessairly increase the log-likelihood function. Nevertheless, it is envisaged that the use of URTSS will provide a more stable algorithm due to its superior performance.

Typically a sequence of likelihood values converge to some stationary points i.e. global (local) maximum or a saddle point. If the sequence of EM is trapped in a saddle point a small random perturbation diverges the algorithm from such a stationary value \cite{McLachlan1997}. In general convergence of the EM algorithm to any stationary points either global (local) maxima or saddle points depends on the initialisation. Here a bounded sequence of the state vectors were used to initialise the algorithm which gives a satisfactory convergence and good state and parameter estimations.

Since the non-linearity is a sigmoid function first-order Taylor linearisation provides a good approximation to the lower bound of the likelihood function, however in general settings with more complicated non-linear behaviours higher order terms might need to be included, this however introduces intractable computational complexity.
\section{Conclusion}
In this paper, a method has been developed to estimate a spatially continuous neural field model from electrophysiological data where the kernel spatial support and the field disturbance characteristics were unknown to the estimator.  It was shown that by choosing the synaptic parameter and the observation noise variance within a plausible range these unknown quantities can be identified. This improves the complexity and the accuracy of the EM algorithm by reducing the number of unknown variables in the estimation process. In addition to the connectivity kernel weights, synaptic parameter, observation noise and disturbance variances were also identified. The empirical distributions of the estimated non-kernel parameters showed small bias with very smalll standard deviations.

>>>>

Here we describe the problem of estimating non-linear state space neural field model using expectation-maximisation (EM) based algorithm \cite{Dempster1977,Shumway2000}. EM is an iterative algorithm that converges on the maximum-likelihood estimates of the parameters. Both the E and the M steps of the EM algorithm increase a lower bound on the log-likelihood function of the parameters given the observed field.  In the E-step we estimate the state $\mathbf x_t$, of the system given the observed field and current estimate of the model parameters to calculate expected log-likelihood. In the M-step we estimate the system parameters by maximising the expected log-likelihood computed in the E-step. The resulting parameters are then used to estimate the distribution of the state vector sequence for the next iteration. The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}.

@@

The estimated supports were then employed in the EM algorithm described in Section \ref{sec:EstimationAlgorithm} to estimate the smoothed states $\hat{\mathbf x}_t^b$, the connectivity kernel weights $\boldsymbol\theta$, the synaptic dynamic $\xi$, and the hyperparameters $\sigma_{\epsilon}^2$ and $\sigma^2_d$. A  square grid of $n_{\theta}=25$ (5 by 5) basis functions with $2.25~mm$ spacing were used to cover the estimated kernel support where $\sigma_{\psi}^2=3.5$. This is shown in Figure.\ref{fig:KernelWidthEstimation} by blue circles, each of which representing full width at half maximum of the connectivity kernel basis functions. The true and estimated connectivity kernel for 100 Monte Carlo simulations are shown in Figure~\ref{fig:KernelEstimates}. Cross-sections of the connectivity kernel estimates for each run are shown in Figure~\ref{fig:KernelEstimateCrossSection}. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel is inside the confidence interval. The large standard deviation is due to the high number of parameters need to be estimated. Figure~\ref{fig:Histogram} depicts the histogram of parameter estimates from 100 simulations where the true parameter value is shown by the solid line and the mean parameter estimate is shown by the dashed line. These histograms indicate small bias in parameter estimates with a very small standard deviation. Examples of neural field reconstruction are given in Figure.~\ref{fig:FieldEstimates}.  The figure shows snapshots at three time instants of the true neural field and the estimated field. In each subplot neural activity is plotted against spatial index in 1D to provide a comparision between true and the estimated underlying fields. Plot of lower bound of the log likelihood versus iterations of the EM is shown in Figure.~\ref{fig:LowerBound}. In general, for the non-linear dynamical systems the actual lower bound can not be claculated analytically, therefore the approximate of the lower bound obtained by URTSS is shown. The change in lower bound drops below 0.009 percent after 15 iterations, confirming
that the algorithm converged to steady parameter values.
\section{Discussion}
The correlation analysis specifies approximate supports for both disturbance covariance function and connectivity kernel. Although the unknown parameters $\xi$ and $\sigma_{\varepsilon}^2$ can be set as described in \S~\ref{subsec:CorrAnal}, the chosen values influence the exact shape of the kernel significantly, the analysis also assumes the sigmoidal firing rate to be in the active region, as a result  the kernel and the disturbance amplitudes cannot be inferred only through correlation analysis. This step might be considered as a prerequisite for the EM algorithm described in \S~\ref{sec:EstimationAlgorithm} where approximated supports are employed to estimate the gain of the kernel and disturbance. Knowing the spatial extent of the kernel greatly increases the speed of the algorithm as the number of basis functions to decompose the kernel reduces significanly compared to the case in which the basis funtions are laid out across the entire spatial domain. This also improves the uncertainty of the estimates because of the lower number of parameters needs to be estimated. On the other hand, the approximated support of the disturbance covariance function facilitates the implementation of the EM algorithm as the disturbance covariance matrix can be factorised into two components, a constant matrix depending only on the inferred support and an unknown scaling parameter. This way the problem of estimating the covariance matrix reduces to estimation of a single scaling parameter.

The additional Taylor expansion step inroduced in the E-step resembles the extended Kalman filtering (EKF) \cite{Jazwinski1970,Chui2009} in which a stationary nonlinear dynamic system is approximated by a linear non-stationary system using the first order Taylor approximation of the nonlinear equations about the state estimates. This may suggest that the extended Kalman smoother is a natural approach to infer the sequence of state vecors in the presented framework. However, extended Kalman smoother (EKS) requires relinearising about the backwards estimates which in turn necessitates the nonlinear mapping to be uniquely invertible. Although this is indeed the case in the neural field equation employed in this work it will impose more computational complexity in the state estimation procedure. In addition, URTSS provides better state approximation compared to the EKS \cite{Haykin2001} and therefore the approximated lower bound will be more accurate  resulting in better parameter estimations.

While EKS has been  applied in the context of EM algorithm for learning a nonlinear model from data \cite{Ghahramani1999} this is the first work (to the best authors' knowlege) where the URTSS has been used to perform state inference in the E-step of the EM algorithm. However, as in the case of EKS since an approximate lower bound is maximised at each time step there is no guarantee that URTSS increases the lower bound on the true likelihood function and therefore the stability of the algorithm cannot be ensured. Nevertheless, the proposed framework is more stable due to better performance of the URTSS.

In almost all applications a sequence of likelihood converges to some stationary values i.e. global (local) maximum or a saddle point. If the sequence of EM is trapped in a saddle point a small random purturbation diverges the EM from the saddle point \cite{McLachlan1997}. In general convergence of the EM algorithm to any stationary point either global (local) maxima or saddle points 
depends on the initialisation. Here a bounded sequence of state vectors were used to initialise the algorithm which gives a satisfactory convergence and good state and good parameter estimations.

Since the non-linearity is a sigmoid function first-order Taylor approximation provides a good approximation to the lower bound of the likelihood function, however in general settings with more complicated nonlinear behaviours higher order terms might need to be included, this however introduces intractable computational complexity.
\begin{itemize}
	\item itterative EM and correlation analysis (maybe)
\end{itemize}
\section{Conclusion}
